### 2021-11-22 chapter2/chapter.tex
proof of properties for the closing operator for the normalised version

%Rather than approximate the sojourn time of \((X(t),\varphi(t)) \in(\calD_k, i)\) by an exponential distribution, we could instead approximate this sojourn time by a phase-type distribution denoted \(\tau_i \sim PH(\bs \alpha, T_{ii}\bs{I}+\bs{S}_i)\) where \(-\bs\alpha \bs{S}^{-1}\bs 1 = \Delta\). That is, the mean of the Phase-type \(PH(\bs \alpha,\bs{S})\) is equal to the cell width, \(\Delta\). The distribution \(PH(\bs \alpha, T_{ii}\bs{I}+\bs{S}_i)\) can be viewed as \(min(U,V)\) where \(U\sim exp(T_{ii})\) and \( V\sim PH(\bs\alpha, \bs{S})\). Thus, expiry of the distribution \(\tau_i\) corresponds to either a change in phase if \(U\) is the minimum, or change in level if \(V\) is the minimum. The latter case we map to a change in level: \(L(t)\) moves from \(\ell\) to \(\ell+1\) if \(c_i>0\), and from \(\ell\) to \(\ell-1\) if \(c_i<0\).

@@

%Let \(\bs{D} = \bs c\bs \alpha\) where \(\bs c = -\bs{C}\bs 1\). Asmussen and Bladt \citep{ab1999} gave such a process a stochastic interpretation (in fact, Asmussen and Bladt worked with rational arrival processes, but I think we need only consider ME renewal processes). Let \(V = span\left\{\alpha, \bs\alpha \bs{S},...,\bs\alpha \bs{S}^{p-1}\right\}\). There exists a vector-valued stochastic process \(\left\{\bs A(t)\right\}_{t}\) with \(\bs A(t)\in V\) for all \(t\) and \(\bs A(0)=\bs\alpha\). Suppose \(\{Y_k\}\), \(k=0,1,...\) are the renewal times, and \(Y_0=0\). Between renewal epochs \(\bs A(t)\) moves deterministically according to \(\bs A(t) = \cfrac{\bs A(Y_{N(t)}) e^{\bs{C}(t-Y_{N(t)})}}{\bs A(Y_{N(t)}) e^{\bs{C}(t-Y_{N(t)})}\bs 1}\). \(\bs A(t)\) jumps at rate \(\bs A(t)\bs{D}\bs 1\) at time \(t\), and upon a jump \(\bs A(Y_k) = \cfrac{\bs A(Y_k-) \bs{D}}{\bs A(Y_k-) \bs{D}\bs 1}=\bs \alpha\) where \(Y_k-\) is the left limit at \(Y_k\). Furthermore, \(\left\{\bs A(t)\right\}\) is a strong Markov process and \(E\left[\bs A(t+s) \mid \bs A(t)\right] = \bs A(t)e^{\bs{Q}t}\) where \(\bs{Q} = \bs{C}+\bs{D}\). We know \(\bs{Q}\bs 1=\bs 0\) and \(dev(\bs{Q})=0\), that is, the dominant eigenvalue of \(\bs{Q}\) is 0.

@@

%Let \(\bs \pi\) be the left eigenvector of \(\bs{Q}\) corresponding to eigenvalue 0. Assume that \(\pi_i\neq0\), \(i=1,...,p\). If this is not the case then construct an invertible matrix \(\bs{V}\) such that \(\bs{V}\bs 1=\bs 1\), then consider the equivalent representation of the ME renewal process with \(ME(\bs\alpha \bs{V}^{-1}, \bs{V}\bs{C}\bs{V}^{-1})\), and pray that \(\pi_i\neq0\), \(i=1,...,p\) (can we show that there is always such a \(\bs{V}\)?) By choosing \(\bs{V}\) such that \(\bs{V}\bs 1=\bs 1\), then we maintain that the right eigenvector corresponding to \(\bs{V}\bs{C}\bs{V}^{-1}\) is still \(\bs 1\). 

@@

The difference between the approximation schemes is the way in which the vector \(\bs a\) is used. We generalise this idea with the concept of \emph{closing operators} which is a linear operator \(V(x):\mathcal A \to \mathbb R,\) for each \(x\in[0,\Delta)\). For example, the closing operator in Equations (\ref{eqn: density approx aug}) is the operator \(V(x),\,x\in[0,\Delta)\) such that for any \(\bs a \in \mathcal A\), 
\begin{align}
	\bs a V(x) &= \bs a e^{\bs Sx}\bs s.\label{eqn: alal}
	%\\\bs a V_-(x-y_\ell) &= \bs a e^{\bs S(x-y_{\ell})}\bs s,\label{eqn:ndaenf}
\end{align}
Similarly, in (\ref{eqn: density approx 3+})
\begin{align}
	\bs a V(x) &= \bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right),\label{eqn:546}
\end{align}
and in (\ref{eqn: density approx 4+})
\begin{align}
	\bs a V(x) &= \bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)}\left[I-e^{S2\Delta}\right]^{-1} \bs s\right).\label{eqn:5462}
\end{align}
The operator \(V(x)\) is a choice which forms part of the definition of the approximation scheme. As we shall see, given certain properties of \(V(x)\), we can prove that the approximation scheme converges, and ensures positivity. In the cases above, all the closing operators (\ref{eqn: alal})-(\ref{eqn:546}) lead to an approximation which converges and, due to their interpretation as probability densities, ensure positivity.

% \begin{rem}
% Since we require \(V\) to be linear, then we actually have 
% \begin{align}
% 	&\int_{\bs a \in\mathcal A}\mathbb P(L(t) = \ell, \bs A(t)\in\wrt \bs a,\phi(t) = j\mid L(0)=\ell_0, \bs A(0) = \bs a_{\ell_0,i}(x_0), \phi(0) = i) \bs a V(x) \nonumber 
% 	\\&= \mathbb E[\bs A(t)1(L(t)=\ell, \phi(t)=j) \mid L(0)=\ell_0, \bs A(0) = \bs a_{\ell_0,i}(x_0), \phi(0) = i]V(x).
% \end{align}
% \end{rem}

% \begin{rem}
	\paragraph{Comments on linearity and normalisation}
	We considered taking the closing operators to be 
	\begin{align}
		\bs a V(x) &= \cfrac{\bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right)}{1-\bs a e^{\bs S 2\Delta}\bs e},\label{eqn: nonlin}
	\end{align}
	which also ensures that the solution always preserves mass (the result of integrating (\ref{eqn: nonlin}) over \([0,\Delta)\) is 1). However, (\ref{eqn: nonlin}) is not a linear operator. The importance of this fact is that for a linear operator, 
	\[\mathbb E[\bs A(t) V(x)] = \mathbb E[\bs A(t)]V(x),\]
	thus, the computation of \(\mathbb E[\bs A(t) V(x)]\) can be achieved by first computing \(\mathbb E[\bs A(t)]\) and then applying \(V(x)\) to the result. In this sense, \(\mathbb E[\bs A(t)]\) contains all of the information about the history of the process up to time \(t\) needed to compute \(\mathbb E[\bs A(t) V(x)]\). This is much the same as in RAPs and QBD-RAPs where \(\mathbb E[\bs A(t)]\) is all that is required to compute probabilities about the future of the processes from time \(t\) onwards.

	When \(V(x)\) is not linear then, in general, 
	\[\mathbb E[\bs A(t) V(x)] \neq \mathbb E[\bs A(t)]V(x).\]
	For example, with (\ref{eqn: nonlin}), 
	\begin{align}
		\mathbb E[\bs A(t) V(x)] = \int_{\bs a \in\mathcal A} \mathbb P(\bs A(t) \in \wrt \bs a) \cfrac{\bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right)}{1-\bs a e^{\bs S 2\Delta}\bs e}.\label{eqn:nonlin2}
	\end{align}
	While the calculation in (\ref{eqn:nonlin2}) is theoretically possible, practically it is not. The seemingly innocuous integral over \(\bs a \in\mathcal A\) actually amounts to an integral over all possible sample paths of the QBD-RAP (i.e.~all possible event times of the QBD-RAP before time \(t\)) and in general, there is no `nice' way to do this computation.

@@

	In practice we use 
	\begin{align}
		\cfrac{\mathbb E\left[\bs A(t)\right] \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right)}{1-\mathbb E\left[\bs A(t)\right] e^{\bs S 2\Delta}\bs e}
	\end{align}
	as a normalised version of (\ref{eqn:546}), and this seems to work well. However, this is not the same as using the closing operator (\ref{eqn: nonlin}).
% \end{rem}

>>>>

%Rather than approximate the sojourn time of \((X(t),\varphi(t)) \in(\calD_k, i)\) by an exponential distribution, we could instead approximate this sojourn time by a phase-type distribution denoted \(\tau_i \sim PH(\bs \alpha, T_{ii}\bs{I}+\bs{S}_i)\) where \(-\bs\alpha \bs{S}^{-1}\bs 1 = \Delta\). That is, the mean of the Phase-type \(PH(\bs \alpha,\bs{S})\) is equal to the cell width, \(\Delta\). The distribution \(PH(\bs \alpha, T_{ii}\bs{I}+\bs{S}_i)\) can be viewed as \(min(U,{\bs v})\) where \(U\sim exp(T_{ii})\) and \( {\bs v}\sim PH(\bs\alpha, \bs{S})\). Thus, expiry of the distribution \(\tau_i\) corresponds to either a change in phase if \(U\) is the minimum, or change in level if \({\bs v}\) is the minimum. The latter case we map to a change in level: \(L(t)\) moves from \(\ell\) to \(\ell+1\) if \(c_i>0\), and from \(\ell\) to \(\ell-1\) if \(c_i<0\).

@@

%Let \(\bs{D} = \bs c\bs \alpha\) where \(\bs c = -\bs{C}\bs 1\). Asmussen and Bladt \citep{ab1999} gave such a process a stochastic interpretation (in fact, Asmussen and Bladt worked with rational arrival processes, but I think we need only consider ME renewal processes). Let \({\bs v} = span\left\{\alpha, \bs\alpha \bs{S},...,\bs\alpha \bs{S}^{p-1}\right\}\). There exists a vector-valued stochastic process \(\left\{\bs A(t)\right\}_{t}\) with \(\bs A(t)\in {\bs v}\) for all \(t\) and \(\bs A(0)=\bs\alpha\). Suppose \(\{Y_k\}\), \(k=0,1,...\) are the renewal times, and \(Y_0=0\). Between renewal epochs \(\bs A(t)\) moves deterministically according to \(\bs A(t) = \cfrac{\bs A(Y_{N(t)}) e^{\bs{C}(t-Y_{N(t)})}}{\bs A(Y_{N(t)}) e^{\bs{C}(t-Y_{N(t)})}\bs 1}\). \(\bs A(t)\) jumps at rate \(\bs A(t)\bs{D}\bs 1\) at time \(t\), and upon a jump \(\bs A(Y_k) = \cfrac{\bs A(Y_k-) \bs{D}}{\bs A(Y_k-) \bs{D}\bs 1}=\bs \alpha\) where \(Y_k-\) is the left limit at \(Y_k\). Furthermore, \(\left\{\bs A(t)\right\}\) is a strong Markov process and \(E\left[\bs A(t+s) \mid \bs A(t)\right] = \bs A(t)e^{\bs{Q}t}\) where \(\bs{Q} = \bs{C}+\bs{D}\). We know \(\bs{Q}\bs 1=\bs 0\) and \(dev(\bs{Q})=0\), that is, the dominant eigenvalue of \(\bs{Q}\) is 0.

@@

%Let \(\bs \pi\) be the left eigenvector of \(\bs{Q}\) corresponding to eigenvalue 0. Assume that \(\pi_i\neq0\), \(i=1,...,p\). If this is not the case then construct an invertible matrix \(\bs{{\bs v}}\) such that \(\bs{{\bs v}}\bs 1=\bs 1\), then consider the equivalent representation of the ME renewal process with \(ME(\bs\alpha \bs{{\bs v}}^{-1}, \bs{{\bs v}}\bs{C}\bs{{\bs v}}^{-1})\), and pray that \(\pi_i\neq0\), \(i=1,...,p\) (can we show that there is always such a \(\bs{{\bs v}}\)?) By choosing \(\bs{{\bs v}}\) such that \(\bs{{\bs v}}\bs 1=\bs 1\), then we maintain that the right eigenvector corresponding to \(\bs{{\bs v}}\bs{C}\bs{{\bs v}}^{-1}\) is still \(\bs 1\). 

@@

The difference between the approximation schemes is the way in which the vector \(\bs a\) is used. We generalise this idea with the concept of \emph{closing operators} which is a linear operator \({\bs v}(x):\mathcal A \to \mathbb R,\) for each \(x\in[0,\Delta)\). For example, the closing operator in Equations (\ref{eqn: density approx aug}) is the operator \({\bs v}(x),\,x\in[0,\Delta)\) such that for any \(\bs a \in \mathcal A\), 
\begin{align}
	\bs a {\bs v}(x) &= \bs a e^{\bs Sx}\bs s.\label{eqn: alal}
	%\\\bs a {\bs v}_-(x-y_\ell) &= \bs a e^{\bs S(x-y_{\ell})}\bs s,\label{eqn:ndaenf}
\end{align}
Similarly, in (\ref{eqn: density approx 3+})
\begin{align}
	\bs a {\bs v}(x) &= \bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right),\label{eqn:546}
\end{align}
and in (\ref{eqn: density approx 4+})
\begin{align}
	\bs a {\bs v}(x) &= \bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)}\right)\left[I-e^{S2\Delta}\right]^{-1} \bs s.\label{eqn:5462}
\end{align}
The operator \({\bs v}(x)\) is a choice which forms part of the definition of the approximation scheme. As we shall see, given certain properties of \({\bs v}(x)\), we can prove that the approximation scheme converges, and ensures positivity. In the cases above, all the closing operators (\ref{eqn: alal})-(\ref{eqn:546}) lead to an approximation which converges and, due to their interpretation as probability densities, ensure positivity.

% \begin{rem}
% Since we require \({\bs v}\) to be linear, then we actually have 
% \begin{align}
% 	&\int_{\bs a \in\mathcal A}\mathbb P(L(t) = \ell, \bs A(t)\in\wrt \bs a,\phi(t) = j\mid L(0)=\ell_0, \bs A(0) = \bs a_{\ell_0,i}(x_0), \phi(0) = i) \bs a {\bs v}(x) \nonumber 
% 	\\&= \mathbb E[\bs A(t)1(L(t)=\ell, \phi(t)=j) \mid L(0)=\ell_0, \bs A(0) = \bs a_{\ell_0,i}(x_0), \phi(0) = i]{\bs v}(x).
% \end{align}
% \end{rem}

% \begin{rem}
	\paragraph{Comments on linearity and normalisation}
	We considered taking the closing operators to be 
	\begin{align}
		\bs a {\bs v}(x) &= \cfrac{\bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right)}{1-\bs a e^{\bs S 2\Delta}\bs e},\label{eqn: nonlin}
	\end{align}
	which also ensures that the solution always preserves mass (the result of integrating (\ref{eqn: nonlin}) over \([0,\Delta)\) is 1). However, (\ref{eqn: nonlin}) is not a linear operator. The importance of this fact is that for a linear operator, 
	\[\mathbb E[\bs A(t) {\bs v}(x)] = \mathbb E[\bs A(t)]{\bs v}(x),\]
	thus, the computation of \(\mathbb E[\bs A(t) {\bs v}(x)]\) can be achieved by first computing \(\mathbb E[\bs A(t)]\) and then applying \({\bs v}(x)\) to the result. In this sense, \(\mathbb E[\bs A(t)]\) contains all of the information about the history of the process up to time \(t\) needed to compute \(\mathbb E[\bs A(t) {\bs v}(x)]\). This is much the same as in RAPs and QBD-RAPs where \(\mathbb E[\bs A(t)]\) is all that is required to compute probabilities about the future of the processes from time \(t\) onwards.

	When \({\bs v}(x)\) is not linear then, in general, 
	\[\mathbb E[\bs A(t) {\bs v}(x)] \neq \mathbb E[\bs A(t)]{\bs v}(x).\]
	For example, with (\ref{eqn: nonlin}), 
	\begin{align}
		\mathbb E[\bs A(t) {\bs v}(x)] = \int_{\bs a \in\mathcal A} \mathbb P(\bs A(t) \in \wrt \bs a) \cfrac{\bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right)}{1-\bs a e^{\bs S 2\Delta}\bs e}.\label{eqn:nonlin2}
	\end{align}
	While the calculation in (\ref{eqn:nonlin2}) is theoretically possible, practically it is not. The seemingly innocuous integral over \(\bs a \in\mathcal A\) actually amounts to an integral over all possible sample paths of the QBD-RAP (i.e.~all possible event times of the QBD-RAP before time \(t\)) and in general, there is no `nice' way to do this computation.

@@

	In practice we may use 
	\begin{align}
		\cfrac{\mathbb E\left[\bs A(t)\right] \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right)}{1-\mathbb E\left[\bs A(t)\right] e^{\bs S 2\Delta}\bs e}
	\end{align}
	as a normalised version of (\ref{eqn:546}), and this seems to work well. However, this is not the same as using the closing operator (\ref{eqn: nonlin}).
% \end{rem}
